---
title: "Basics_of_R"
output: html_document
date: "2022-07-13"
---

The assignment operator in R is <- symbol. 
```{r}
x<-5
print(x)
```

## Creating sequences in R
to create sequence in R. we use the : symbol


```{r}
y<-1:20
print(y)
```


## Different types of classes in R. 

Everything that we encounter in R is called as an object. now each object has a class. 
the 5 basics classes in R is
1) Character
2) Numeric
3) Integer
4) Complex 
5) Logical 

The most basic object in R is a vector. 

a vector is a collection of objects from the same class. 
To store objects from different classes we use lists. 
```{r}
c1<-'a'
n1<-10.6
n2<- 10L
co1<- 5 + 4i
L1<- TRUE
temp<- as.numeric(L1)
L2<- list(1,'a',T,5+4i)
print(c1)
print(n1)
print(n2)
print(co1)
print(L1)
print(L2)
print(temp)
```


## Numbers 

Numbers in R is treated as numeric objects. (double precision type)
if you specifically want integers. then L should be a suffix at the number.
INF - represents infinity. it is a special type of numeric object. 
similarly Nan is also special type of numeric object. Nan represents undefined. ex 0/0. 
To convert a datatype of an object. we use 
1) as.numeric()
2) as.logical()
3) as.character()


list are used to store objects of different class. 
we create a list using the list keyword.





