---
title: "Basics_of_R"
output: html_document
date: "2022-07-13"
---

The assignment operator in R is <- symbol. 
```{r}
x<-5
print(x)
```

## Creating sequences in R
to create sequence in R. we use the : symbol


```{r}
y<-1:20
print(y)
```


## Different types of classes in R. 

Everything that we encounter in R is called as an object. now each object has a class. 
the 5 basics classes in R is
1) Character
2) Numeric
3) Integer
4) Complex 
5) Logical 

The most basic object in R is a vector. 

a vector is a collection of objects from the same class. 
To store objects from different classes we use lists. 
```{r}
c1<-'a'
n1<-10.6
n2<- 10L
co1<- 5 + 4i
L1<- TRUE
temp<- as.numeric(L1)
L2<- list(1,'a',T,5+4i)
print(c1)
print(n1)
print(n2)
print(co1)
print(L1)
print(L2)
print(temp)
```


## Numbers 

Numbers in R is treated as numeric objects. (double precision type)
if you specifically want integers. then L should be a suffix at the number.
INF - represents infinity. it is a special type of numeric object. 
similarly Nan is also special type of numeric object. Nan represents undefined. ex 0/0. 
To convert a datatype of an object. we use 
1) as.numeric()
2) as.logical()
3) as.character()


list are used to store objects of different class. 
we create a list using the list keyword.



## Matrices 


They are special vector in R with the dimension attribute. The dimension attribute is an integer vector of size 2. (Ncol,Nrow)
We create a matrix using the keyword matrix. 
It has 2 attributes 
- [ ] Dims - returns the number of dimensions 
- [ ] Attributes 
Matrix is created column wise. 
So first all the columns are filled, that is all the rows in that column are filled first. 

Matrices are also created using a vector. 
Ex x<-1:10
So x is sequence of 1 to 10 
Then dim(x)<- c(2,5)
So we have assigned the dimensions of x as 5 columns and 2 rows. 

Another way of creating a matrix is by row binding and column binding. 

So in Cbind the columns are filled first. So x fills the first column and y fills the second column. 
So in Rbind the rows are filled first. 


```{r}
#creating a matrix 
x<- matrix(ncol = 2, nrow = 2)
# creating a matrix using vectors
y<- 1:10
# we create a matrix by creating a vector containing the dimensions and assigning the vector as the 
# dimensions of sequence y.
dim(y)=c(2,5)
print(y)
x<- 1:3
y<- 10:12
cbind(x,y)
rbind(x,y)
```


## Factors 



Factors are special types of vectors that are used to represent categorical data. 
There are 2 types of factors (nominal data, ordinal data)
- [ ] Unordered 
- [ ] Ordered 

One can think of factor as an integer vector where each integer has a label. 
A factor can be created using the keyword factor. 
X<- factor(c(‘yes’,’no’,’yes’))
Here are there are two types of distinct values (‘yes’, ‘no’)
These are called as levels.

Levels are the attribute for factor. 
Unclass function will strip the factor from its class and represent it as the integer vector. 

The order of the levels can be set using the levels argument to factor (). This can be important for linear modelling. 

So in the example of yes and no. Alphabetically no should be level 1 and yes should be 2. 
We can change it using the levels parameter. 

X<- factor(c(‘yes’,’no’,’yes’), levels = c(‘yes’, ’no’))

```{r}
# Creating a factor in R, we use the factor keyword
X<- factor(c("yes","yes","no"))
print(X)
# Unclass strips the class from the factor and then converts into a integer vector
unclass(x)
# In order to assign levels we pass the levels parameter to the factor and manually assign the functions
X<- factor(c("yes","yes","no"), levels = c('yes','no'))
print(X)
```


## Missing values in R

They are denoted by either na or nan 
There 2 functions to check for na and nan values 
- [ ] Is.na
- [ ] Is.nan 
Na values have a class also, so there are integer na and character na etc. 
nan value is also na but converse is not true. 
is.na value is return True if nan values are present but is.nan will not return true if na is present. 
```{r}
x<-c(1,2,NA,10,3)
is.na(x)
x<-c(1,2,NA,NaN,10,3)
is.nan(x)
is.na(x)
```

## Data frames

Data frames are used to store tabular data. 
- [ ] They are special type of list where every elements must be of the same type 
- [ ] Each element of the list can be thought as a column and length of element can be thought as the number of rows
- [ ] It can store objects of different classes 
- [ ] They have special attribute called row.names 
- [ ] They are created by calling read.table or read.csv()
- [ ] Can be converted to a matrix by calling data.matrix()
- [ ] They can also be created using the data.frame function 
 
X<- data.frame(foo = 1:4, bar = c(T,T,F,F))
nrow returns the total number of rows 
ncol returns the total number of columns 

R objects can have names, now this is not just true for data frames but all R objects. 
X<-1:3
Names(x) will return null 
We can now assign names to the object x. 
Names(x) <- c(‘foo’, ‘bar’, ‘norf’)
X
Names(x) 
matrices can have names but to assign the names we must pass the list which contains a list of row names and list of column names
we use the dimnames function 
dimnames(m) <- list(c('a','b'),c('c','d'))
the above line works with 2*2 matrix.
```{r}
# using the data.frame we can create a dataframe. here foo and bar are the column names. 
X<- data.frame(foo = 1:4, bar = c(T,T,F,F))
X
nrow(X)
ncol(X)
X<-1:3
names(X)
names(X)<- c('foo','bar','norf')
X
names(X)
```
## Reading and writing data in R 

- [ ] Read.table, read.csv for reading tabular data 
- [ ] ReadLines for reading lines of a text file 
- [ ] Source for reading in R code files (inverse of dump)
- [ ] Dget, for reading R code files (inverse of Dput)
- [ ] Load, for reading saved workspaces 
- [ ] Unserialize, for single R objects in Binary form 

The read.table function is one of the most commonly used functions for reading data. It has few important arguments. 
- [ ] File name - name of the files 
- [ ] Header - indicates if the file has a header line 
- [ ] Sep  - a string indicating how the columns are separated 
- [ ] ColClasses -  a character vector indicating the class of the column 
- [ ] Nrow - number of rows in the dataset 
- [ ] Comment.char - character string indicating the comment character 
- [ ] Skip - number of lines to skip from beginning
- [ ] StringAsFactors - should character variable be coded as factors


## Textual format 

Dumping and dputing are useful because the resulting textual format is editable and in case of corruption potentially recoverable. 

Unlike writing out a table or csv file, dump and dput  preserve the metadata. 
So that another user doesn’t have to specify it all over again. 

Textual format works better with version control system. 

If y is the data frame 
Dput(y,file = “y.R”)
Now retrieve the file, we use the dget function 
New.y <- dget(‘y.R’)

Dump is similar to Dput however it works with multiple objects. 
X<- “foo”
Y<- data.frame(a=1,b=“a”)
Dump(c(“X”, “Y”, file = “data.R”)
Rm(x,y)
Now to retrieve the files in R, we use the source function. 
Source(‘data.R’)

Interfaces to the outside world

These are functions that connect to the outside world. 
Most of connections can be made to files 
- [ ] File - opens a connection to a file 
- [ ] Gzfile - opens a connection to a compressed file with gz2 
- [ ] Bzfile - opens a connection to a file compressed with bz2
- [ ] URL - opens a connection to a web page 

File function has a few arguments. 
1. description is the name of the file 
2. Open is a code indicating
3.  R read only 
4. W writing.(it opens a new file)
5. A - appending 
6. Rb,wb,ab are reading, writing, appending in binary mode (windows)


To close the connections we use the close function.
ReadLines are useful for reading lines from a web page

Con <- url(link)

X<- readLines(con)



## Sub setting 

There are couple options that are used for sub setting in R 
- [ ] [ always returns an object of the same class as the original; can be used select more than one element. If you subset a list you will get back a list. 
- [ ] [[ is used to extract elements of a list or a data frame; it can only be used to extract a single element and class of the returned object will not necessarily be a list or a data frame. 
- [ ] $ is used to extract elements of a list or data frame by name; semantics are similar to [[

Ex x = c(“a”,”b”,”c”,”d”,”a”)
X[1] will return a 
X[1:4] will return the first 4 elements 
X[x>a] will return all the elements that are greater than the letter “a”

The above is an example of vector sub setting. 

List sub setting 
X<- list(foo = 1:4, bar =0.6)
X[1] - single bracket will return the object of the same class so since we are sub setting a list, the output is also a list. 
X[1] - $foo - name of the list 
1 2 3 4 
X[[1]] - double bracket will returns objects of different class 
1 2 3 4
So in the first case it returned a list 
2nd case it returned a sequence 

X$bar 
0.6 
We subsetted the list using the name of the object “bar”

To extract multiple elements of the list we use the [. We cannot extract multiple elements using the [[ or $ symbol 
X[c(1,3)] 
$foo 
1,2,3

Another way of subsetting is 

Name <- “foo”
X[[name]] # it computed the index with the name foo 

To subset list within a list 

We can do it using 2 methods 

X[c(1,3)]
So this is will return the 1st of the list x and 3 rd element of the extracted list. 

X[[1]][[3]]
Using double brackets 


Subsetting matrices 

X<- matrix(1:6,2,4)
So the first parameter is the row and second parameter is the column. 
X[1,2]
So X is 2*4 matrix 
So we are trying to find the 1 row and the 2nd column. 
X[1,]
It will return the entire 1st row 
X[,2]
It will return the entire 2nd column. 

So in this case [ will not return an object of the same class. To avoid this behaviour we use a parameter called drop and set it to FALSE.

subsetting a dataframe

dataframes are divided into [rows,columns]
so if we want our output to be a dataframe 
then we need to write our condition and then end it with a ","
if comma is not mentioned then the output is a sequence. 



## Partial matching 

X<- list(‘Anything ’ = 1:5)
So if typing Anything  is difficult everything time we can use partial matching. 
X[[‘A’]] - will return NULL 
But 
X[[‘A’,exact = FALSE]]
It will return the output of the object named anything that is 
1 2 3 4 5 


## Removing na values. 

X<- c(1,2,4,Na,Na,5)
Bad <- is.na(X)
X[!bad]
!bad is all the values that are not null. 

So if 2 vectors have missing values in the same positions then we can use 
Complete.cases function 

Complete.cases(x,y) 
Will return a logical vector that stores true if the value is not Na and false if it is. 
Note == is used to equality 

There 2 methods for matrix multiplications. 
- [ ] Element wise. So we use x*y
- [ ] True matrix multiplication; x%%y is used. 


## Control Structures

for loop is simple loop that runs for a certain number of times. here the number of times is defined. 

while loop is loop that runs until a certain condition is false. we need to be careful, as while loops may run infinite number of times if our condition is not defined. 

repeat is special loop that initiates an infinite loop. the only way to exit the loop is to call break.

the next statement is similar to continue statement in other programming language. so if next is encountered then R will skip an interation and move to the next iteration. 
```{r}
x<- 5
if(x<5){
  y<-10
}else {
  y<-0
}
print(y)
#There is another way of writing the above code 
y<-if(x<6){
  10
}else{
  0
}
print(y)

# Loops
# Looping through a sequence
for(i in 1:10){
  print(i)
}
# loop through a character vector 
for(c in c('a','b','c','d')){
  print(c)
}
t<- 1:10
# seq_along function will create another vector containing values from 1 till length of the function
for(i in seq_along(t)){
  print(i)
}
m<- matrix(1:20,4,5)
# seq_len function converts an integer into a seqence. if i= 2 then the sequence created is 1:2
for(i in seq_len(nrow(m))){
  for(j in seq_len(ncol(m))){
    print(m[i,j])
  }
}
#While loop syntax 
count<- 0 
while(count<10){
print(count)
count<- count+1
}
count<-0
t<-10
# to run multiple conditions in a while loop we use && and ||. they are "and" and or repectively. 
while(count<10 && t>0){
print(count)
count<- count+1
t<- t-2
}
# repeat loop
print(t)
repeat {
  if (t==10){
    t<-t+1
    next # it will skip the iteration when t = 10
  }else if (t%%2==1){# the modulo is symbol in R is %%
      print('Leaving the loop')
    break
  }
  t<-t+2
}
```



