---
title: "Basics_of_R"
output: html_document
date: "2022-07-13"
---

The assignment operator in R is <- symbol. 
```{r}
x<-5
print(x)
```

## Creating sequences in R
to create sequence in R. we use the : symbol


```{r}
y<-1:20
print(y)
```


## Different types of classes in R. 

Everything that we encounter in R is called as an object. now each object has a class. 
the 5 basics classes in R is
1) Character
2) Numeric
3) Integer
4) Complex 
5) Logical 

The most basic object in R is a vector. 

a vector is a collection of objects from the same class. 
To store objects from different classes we use lists. 
```{r}
c1<-'a'
n1<-10.6
n2<- 10L
co1<- 5 + 4i
L1<- TRUE
temp<- as.numeric(L1)
L2<- list(1,'a',T,5+4i)
print(c1)
print(n1)
print(n2)
print(co1)
print(L1)
print(L2)
print(temp)
```


## Numbers 

Numbers in R is treated as numeric objects. (double precision type)
if you specifically want integers. then L should be a suffix at the number.
INF - represents infinity. it is a special type of numeric object. 
similarly Nan is also special type of numeric object. Nan represents undefined. ex 0/0. 
To convert a datatype of an object. we use 
1) as.numeric()
2) as.logical()
3) as.character()


list are used to store objects of different class. 
we create a list using the list keyword.



## Matrices 


They are special vector in R with the dimension attribute. The dimension attribute is an integer vector of size 2. (Ncol,Nrow)
We create a matrix using the keyword matrix. 
It has 2 attributes 
- [ ] Dims - returns the number of dimensions 
- [ ] Attributes 
Matrix is created column wise. 
So first all the columns are filled, that is all the rows in that column are filled first. 

Matrices are also created using a vector. 
Ex x<-1:10
So x is sequence of 1 to 10 
Then dim(x)<- c(2,5)
So we have assigned the dimensions of x as 5 columns and 2 rows. 

Another way of creating a matrix is by row binding and column binding. 

So in Cbind the columns are filled first. So x fills the first column and y fills the second column. 
So in Rbind the rows are filled first. 


```{r}
#creating a matrix 
x<- matrix(ncol = 2, nrow = 2)
# creating a matrix using vectors
y<- 1:10
# we create a matrix by creating a vector containing the dimensions and assigning the vector as the 
# dimensions of sequence y.
dim(y)=c(2,5)
print(y)
x<- 1:3
y<- 10:12
cbind(x,y)
rbind(x,y)
```


## Factors 



Factors are special types of vectors that are used to represent categorical data. 
There are 2 types of factors (nominal data, ordinal data)
- [ ] Unordered 
- [ ] Ordered 

One can think of factor as an integer vector where each integer has a label. 
A factor can be created using the keyword factor. 
X<- factor(c(‘yes’,’no’,’yes’))
Here are there are two types of distinct values (‘yes’, ‘no’)
These are called as levels.

Levels are the attribute for factor. 
Unclass function will strip the factor from its class and represent it as the integer vector. 

The order of the levels can be set using the levels argument to factor (). This can be important for linear modelling. 

So in the example of yes and no. Alphabetically no should be level 1 and yes should be 2. 
We can change it using the levels parameter. 

X<- factor(c(‘yes’,’no’,’yes’), levels = c(‘yes’, ’no’))

```{r}
# Creating a factor in R, we use the factor keyword
X<- factor(c("yes","yes","no"))
print(X)
# Unclass strips the class from the factor and then converts into a integer vector
unclass(x)
# In order to assign levels we pass the levels parameter to the factor and manually assign the functions
X<- factor(c("yes","yes","no"), levels = c('yes','no'))
print(X)
```


## Missing values in R

They are denoted by either na or nan 
There 2 functions to check for na and nan values 
- [ ] Is.na
- [ ] Is.nan 
Na values have a class also, so there are integer na and character na etc. 
nan value is also na but converse is not true. 
is.na value is return True if nan values are present but is.nan will not return true if na is present. 
```{r}
x<-c(1,2,NA,10,3)
is.na(x)
x<-c(1,2,NA,NaN,10,3)
is.nan(x)
is.na(x)
```

## Data frames

Data frames are used to store tabular data. 
- [ ] They are special type of list where every elements must be of the same type 
- [ ] Each element of the list can be thought as a column and length of element can be thought as the number of rows
- [ ] It can store objects of different classes 
- [ ] They have special attribute called row.names 
- [ ] They are created by calling read.table or read.csv()
- [ ] Can be converted to a matrix by calling data.matrix()
- [ ] They can also be created using the data.frame function 
 
X<- data.frame(foo = 1:4, bar = c(T,T,F,F))
nrow returns the total number of rows 
ncol returns the total number of columns 

R objects can have names, now this is not just true for data frames but all R objects. 
X<-1:3
Names(x) will return null 
We can now assign names to the object x. 
Names(x) <- c(‘foo’, ‘bar’, ‘norf’)
X
Names(x) 
matrices can have names but to assign the names we must pass the list which contains a list of row names and list of column names
we use the dimnames function 
dimnames(m) <- list(c('a','b'),c('c','d'))
the above line works with 2*2 matrix.
```{r}
# using the data.frame we can create a dataframe. here foo and bar are the column names. 
X<- data.frame(foo = 1:4, bar = c(T,T,F,F))
X
nrow(X)
ncol(X)
X<-1:3
names(X)
names(X)<- c('foo','bar','norf')
X
names(X)
```

